---
title: Exploration of Loan Data from Prosper by Yajie ZHU
output: 
  html_document: 
    css: ~/templates/styles/basic.css
    fig_caption: yes
    number_sections: yes
---

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(ggplot2)
library(tidyverse)
library(forcats)
writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
dat <- read_csv('data/prosperLoanData.csv')
```

The dataset in this report contains 113,937 loans with 81 variables on 
each loan, including loan amount, borrower rate (or interest rate), 
current loan status, borrower income, borrower employment status, 
borrower credit history, and the latest payment information.

# Data Checking Section

## Check Repeated Records

There are `r nrow(dat)` rows in this dataset while the number of the unique 
value of `ListingKey` is only `r length(unique(dat$ListingKey))` (less than 
the number of the rows), which should be checked before further analysis.

So I checked the count of each ListingKey and show the top 5 ListingKeys below 
which appear more than once in the dataset:

```{r check-ListingKey}
count(dat, ListingKey, sort = T) %>% filter(n > 1) %>% head(5)
```

For the ListingKey ("17A93590655669644DB4C06") which appears six times in the
dataset, I checked all the six records and found that the `ProsperScore` column
is different among these records. However, this is the only difference exists 
in the data and I think there should be an addtional column which indicates when
the `ProsperScore` was changed.

```{r results='hide'}
filter(dat, ListingKey == "17A93590655669644DB4C06") %>% unique()
```

Anyway, there are no repeated records for ListingKey ("17A93590655669644DB4C06")
so no further action is necessary to be taken. Furthermore, after run the code
`unique(dat) %>% dim()` it seems that there are no repeated records in the whole
dataset.

```{r, results='hide'}
unique(dat) %>% dim()
```

## Check Missing Data

It's an important step to check the missing value in the dataset and decide 
whether the missing data will affect the analysis significantly before 
performing any analysis. So I calculated the ratio of missing value for 
each column of this dataset and printed the columns of which the missing data 
is more than 20% in the below table:

```{r}
na_ratio <- map(dat, ~mean(is.na(.))) %>% as.matrix()
na_ratio <- tibble(col = rownames(na_ratio),
                   ratio = as.numeric(na_ratio))
na_ratio %>% arrange(-ratio) %>% filter(ratio > 0.2)
```

At first it looks like that there are 18 variables whose missing data is more 
than 20%, however, after reading the data variables definitions, it turns out 
the `NA` value for each of these variables has its specific meaning, which 
means they are not really the "missing value". Therefore, I believe that this 
is probably a clean dataset and it's time to play with the data!

# Univariate Plots Section

## Ask the Questions

For the Univariate Plots Section, I choose the following questions to explore
in this project:

1. What's the date period of this dataset?
1. What's the distribution of the category of the listing that 
the borrower selected when posting their listing?
1. What's the percentage of each LoanStatus in this dataset?
1. What's the percentage of each EmploymentStatus in this dataset?
1. What's the distribution of EmploymentStatusDuration in this dataset?
1. What's the percentage of the Homeowner in the Borrowers?
1. What's the percentage of the defaulted loans and what's the distribution
of the cycle of these defaulted loans?
1. What's the distribution of ProsperScore in this dataset?

## What's the date period of this dataset?

At the first of all, I would like to check the time period of this dataset:

```{r}
summary(dat$ListingCreationDate)
```

The time period of this dataset is from 2005-11-09 to 2014-03-10.

## What's the distribution of the category of the listing?

```{r}
dat %>%
  mutate(`ListingCategory (numeric)` = 
           factor(`ListingCategory (numeric)`,
                  levels = unique(dat$`ListingCategory (numeric)`))) %>%
  mutate(ListingCategory = fct_recode(`ListingCategory (numeric)`,
    "Not Available" = "0", 
    "Debt Consolidation" = "1", 
    "Home Improvement" = "2",
    "Business" = "3",
    "Personal Loan" = "4",
    "Student Use" = "5",
    "Auto" = "6",
    "Other" = "7",
    "Baby&Adoption" = "8",
    "Boat" = "9",
    "Cosmetic Procedure" = "10",
    "Engagement Ring" = "11",
    "Green Loans" = "12",
    "Household Expenses" = "13",
    "Large Purchases" = "14",
    "Medical/Dental" = "15",
    "Motorcycle" = "16",
    "RV" = "17",
    "Taxes" = "18",
    "Vacation" = "19",
    "Wedding Loans" = "20") %>% fct_infreq() %>% fct_rev()) %>%
    ggplot(aes(ListingCategory)) + geom_bar() + coord_flip()
```

It's interesting to see that the most popular ListingCategory is 
"Debt Consolidation", which looks like morn than the sum of all the other types.
The second and third popular ListingCategory are "Not Available" and "Other",
which provides little effective information about the usage of the debt.

## What's the percentage of each LoanStatus?

```{r echo=FALSE}
dat %>% 
  mutate(LoanStatus = LoanStatus %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(LoanStatus)) + geom_bar() + coord_flip()
```

We can see that only a little fraction of the loans are in the "Past Due" 
status, and the longer the past due days, the less numbers they are. 

However, the number of the "Chargedoff" loans is 
`r sum(dat$LoanStatus == 'Chargedoff')` and the number of the "Defaulted" 
loans is `r sum(dat$LoanStatus == 'Defaulted')`, which means the default risk
is high and should be addressed by the investors (Loans of Chargedoff status 
and Defaulted status stand for `r paste0(round(100*mean(dat$LoanStatus %in% c("Chargedoff", "Defaulted")), 2), "%")` of the total loans).

## What's the percentage of each EmploymentStatus?

```{r}
dat %>%
  mutate(EmploymentStatus = EmploymentStatus %>% 
           fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(EmploymentStatus)) + 
  geom_bar() + coord_flip()
```

There is no surprise that the majority of the borrowers have a job, because 
people with no job have difficulties to obtian the loans.

## What's the distribution of EmploymentStatusDuration?

```{r}
dat %>%
  ggplot(aes(EmploymentStatusDuration)) +
  geom_histogram(binwidth = 10, fill = "blue", 
                 color = "black", alpha = 0.6)
```

```{r}
summary(dat$EmploymentStatusDuration)
```

The figure above shows that the distribution of EmploymentStatusDuration is 
positive skewed, and the summary of EmploymentStatusDuration shows that about
one quarter have more than 10 years of employment, and about one half have more
than 5 years of employment.

## What's the percentage of the Homeowner in the Borrowers?

```{r}
dat %>%
  ggplot(aes(IsBorrowerHomeowner)) +
  geom_bar(width = 0.5, fill = "blue", alpha = 0.8)
```

It shows that the percentage of the homeowner in the borrowers is about one half.

## What's the percentage of the defaulted loans?

The data variable definitions show that for the "LoanFirstDefaultedCycleNumber"
column: If the loan has not charged off the value will be null. So we can use
this variable to compute the percentage of the defaulted loans, which is
`r paste0(round(100*mean(!is.na(dat$LoanFirstDefaultedCycleNumber)), 2), "%")`.

## What's the distribution of the cycle of these defaulted loans?

Furthermore, the distribution of the cycle of these defaulted loans may reveal
more information:

```{r}
dat %>%
  filter(!is.na(LoanFirstDefaultedCycleNumber)) %>%
  ggplot(aes(LoanFirstDefaultedCycleNumber)) +
  geom_bar(fill = "blue", alpha = 0.8)
```

It's interesting to find that if default didn't occur in the first two cycles, 
then it's unlikely to occur in the following three cycles.

## What's the distribution of ProsperScore?

```{r}
dat %>%
  filter(!is.na(ProsperScore)) %>%
  ggplot(aes(ProsperScore)) +
  geom_bar(fill = "blue", alpha = 0.8)
```

The distribution of ProsperScore looks like a normal distribution while the top
3 most frequently ProsperScores are 4, 6, and 8, and the top 2 least frequently 
ProsperScores are 1 and 11. However, the data varibale definitions show that the
ProsperScore ranges from 1-10, with 10 being the best, or lowest risk score. So
I decide to double check the ProsperScore field:

```{r}
count(dat, ProsperScore) %>% filter(ProsperScore == 11)
```

**It seems like that there are 1456 rows of which the ProsperScore equals 11, but
11 is not a valid score based on the data varibale definitions. So I suspect 
that this maybe a data error or something else, which need to be investigated
in the further analysis.**

# Univariate Analysis

## What is the structure of your dataset?

There are 113,937 loans with 81 variables including loan amount, 
borrower rate (or interest rate), current loan status, 
borrower income, borrower employment status, borrower credit history, 
and the latest payment information in this dataset.

There are so many variables in this dataset so it is difficult to explore each 
variable one by one. Instead, it woule be more appropriate to adopt a 
problem-oriented approach to explore this dataset. Actually I used this approach
in this project.

## What is/are the main feature(s) of interest in your dataset?

The main features of interest in this dataset are:

1. `LoanFirstDefaultedCycleNumber`;
1. `ProsperScore`, `ProsperRating (Alpha)`, `ProsperRating (numeric)`;
1. `CreditScoreRangeLower`, `CreditScoreRangeUpper`.

I choose these features as the main features because for the loan business, 
the most important concern is the credit/default risk. For the creditor/lender,
what they most want to forecast is whether the borrower will default on the loan
and when the borrower will most likely default.

## What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Of the six main features of interest described above, I choose 
`LoanFirstDefaultedCycleNumber` as the target feature, and try to use 
`ProsperScore`, `ProsperRating (Alpha)`, `ProsperRating (numeric)`, 
`CreditScoreRangeLower`, `CreditScoreRangeUpper` as well as other features to
predict whether the borrower will default and when the borrower will most 
likely default.

## Did you create any new variables from existing variables in the dataset?

For the Univariate Plots Section, I didn't create new variables from existing
variables, but I did modify some variables (such as changing the factor levels
or reorder the factor) in order to make better visualisations.

However, for the Bivariate Plots (Analysis) Section and 
Multivariate Plots (Analysis) Section, I will create some new variables to get
better understanding of the data and build better models.

## Of the features you investigated, were there any unusual distributions?

Of the features you investigated EmploymentStatusDuration and 
LoanFirstDefaultedCycleNumber don't follow normal distribution and both of them 
are positive skewed.

**Did you perform any operations on the data to tidy, adjust, or change the form
of the data? If so, why did you do this?**

As I mentioned before, I modify some variables (such as changing the factor levels
or reorder the factor) in order to make better visualisations.

# Bivariate Plots Section

1. What's the distribution of IncomeRange of the Borrowers (only for 
the group have verifiable income)? 
1. What's the distribution of StatedMonthlyIncome of the Borrowers (only for 
the group have verifiable income)?

```{r}
dat %>%
  ggplot(aes(IncomeRange)) +
  geom_bar() +
  facet_wrap(~IncomeVerifiable) +
  coord_flip()
```


```{r}
dat %>%
  ggplot(aes(StatedMonthlyIncome)) +
  geom_histogram(binwidth = 100) +
  facet_wrap(~IncomeVerifiable) +
  coord_flip()
```


> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r}
count(dat, `ProsperRating (numeric)`, `ProsperRating (Alpha)`)
```


```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!
