---
title: "Exploration of the Loan Data from Prosper"
author: "Yajie ZHU"
date: "`r Sys.Date()`"
output: 
  html_document: 
    css: css/basic.css
    number_sections: yes
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
	echo = FALSE, cache = TRUE, message = FALSE, warning = FALSE,
	fig.asp = 0.618, fig.align = "center", out.width = "80%"
)
library(tidyverse)
library(magrittr)
library(forcats)
library(maps)
library(e1071)
library(AUC)
writeLines(capture.output(sessionInfo()), "sessionInfo.txt")
```

The dataset in this report contains 113,937 loans with 81 variables on each
loan, including loan amount, borrower rate (or interest rate), current loan
status, borrower income, borrower employment status, borrower credit history,
and the latest payment information.

```{r Load-the-Data}
# When import csv files into R data.frame, it's important to check whether each
# column has been converted to the correct data type. In order to make this 
# process more consistent and robust, it's the best practice to assgin the 
# col_types param for every column one by one. Even though the code will be 
# long and uncomfortable to read, however, it's worth to do this.
dat <- readr::read_csv(
	'data/prosperLoanData.csv.zip',
	col_types = cols(
		ListingKey = col_character(),
		# parse_guess: col_integer()
		ListingNumber = col_character(),
		ListingCreationDate = col_datetime(format = ""),
		# parse_guess: col_charater()
		CreditGrade = 
			col_factor(c("NC", "HR", "E", "D", "C", "B", "A", "AA")),
		Term = col_integer(),
		# parse_guess: col_character()
		LoanStatus = col_factor(levels = NULL),
		ClosedDate = col_datetime(format = ""),
		BorrowerAPR = col_double(),
		BorrowerRate = col_double(),
		LenderYield = col_double(),
		EstimatedEffectiveYield = col_double(),
		EstimatedLoss = col_double(),
		EstimatedReturn = col_double(),
		# parse_guess: col_integer()
		`ProsperRating (numeric)` = 
			col_factor(levels = c(1, 2, 3, 4, 5, 6, 7)),
		# parse_guess: col_character()
		`ProsperRating (Alpha)` = 
			col_factor(levels = c("HR", "E", "D", "C", "B", "A", "AA")),
		ProsperScore = col_double(),
		# parse_guess: col_integer()
		`ListingCategory (numeric)` = col_factor(levels = NULL),
		# parse_guess: col_character()
		BorrowerState = col_factor(levels = NULL),
		# parse_guess: col_character()
		Occupation = col_factor(levels = NULL),
		# parse_guess: col_character()
		EmploymentStatus = col_factor(levels = NULL),
		EmploymentStatusDuration = col_integer(),
		# parse_guess: col_character()
		IsBorrowerHomeowner = col_factor(levels = NULL),
		# parse_guess: col_character()
		CurrentlyInGroup = col_factor(levels = NULL),
		GroupKey = col_character(),
		DateCreditPulled = col_datetime(format = ""),
		CreditScoreRangeLower = col_integer(),
		CreditScoreRangeUpper = col_integer(),
		FirstRecordedCreditLine = col_datetime(format = ""),
		CurrentCreditLines = col_integer(),
		OpenCreditLines = col_integer(),
		TotalCreditLinespast7years = col_integer(),
		OpenRevolvingAccounts = col_integer(),
		OpenRevolvingMonthlyPayment = col_double(),
		InquiriesLast6Months = col_integer(),
		TotalInquiries = col_double(),
		CurrentDelinquencies = col_integer(),
		AmountDelinquent = col_double(),
		DelinquenciesLast7Years = col_integer(),
		PublicRecordsLast10Years = col_integer(),
		PublicRecordsLast12Months = col_integer(),
		RevolvingCreditBalance = col_double(),
		BankcardUtilization = col_double(),
		AvailableBankcardCredit = col_double(),
		TotalTrades = col_double(),
		`TradesNeverDelinquent (percentage)` = col_double(),
		TradesOpenedLast6Months = col_double(),
		DebtToIncomeRatio = col_double(),
		# parse_guess: col_character()
		IncomeRange = col_factor(levels = 
			c("$0", "$1-24,999", "$25,000-49,999", "$50,000-74,999",
				"$75,000-99,999", "$100,000+", "Not employed", "Not displayed")),
		# parse_guess: col_character()
		IncomeVerifiable = col_factor(levels = NULL),
		StatedMonthlyIncome = col_double(),
		LoanKey = col_character(),
		TotalProsperLoans = col_integer(),
		TotalProsperPaymentsBilled = col_integer(),
		OnTimeProsperPayments = col_integer(),
		ProsperPaymentsLessThanOneMonthLate = col_integer(),
		ProsperPaymentsOneMonthPlusLate = col_integer(),
		ProsperPrincipalBorrowed = col_double(),
		ProsperPrincipalOutstanding = col_double(),
		ScorexChangeAtTimeOfListing = col_integer(),
		LoanCurrentDaysDelinquent = col_integer(),
		LoanFirstDefaultedCycleNumber = col_integer(),
		LoanMonthsSinceOrigination = col_integer(),
		# parse_guess: col_integer()
		LoanNumber = col_character(),
		LoanOriginalAmount = col_integer(),
		LoanOriginationDate = col_datetime(format = ""),
		LoanOriginationQuarter = col_character(),
		MemberKey = col_character(),
		MonthlyLoanPayment = col_double(),
		LP_CustomerPayments = col_double(),
		LP_CustomerPrincipalPayments = col_double(),
		LP_InterestandFees = col_double(),
		LP_ServiceFees = col_double(),
		LP_CollectionFees = col_double(),
		LP_GrossPrincipalLoss = col_double(),
		LP_NetPrincipalLoss = col_double(),
		LP_NonPrincipalRecoverypayments = col_double(),
		PercentFunded = col_double(),
		Recommendations = col_integer(),
		InvestmentFromFriendsCount = col_integer(),
		InvestmentFromFriendsAmount = col_double(),
		Investors = col_integer()
	)
)
```

# Data Audit Section

## Check Repeated Records

There are `r nrow(dat)` rows in this dataset while the number of the unique
value of `ListingKey` is only `r length(unique(dat$ListingKey))` (less than the
number of the rows), which should be checked before further analysis.

So I checked the count of each ListingKey and the top 5 ListingKeys which appear
more than once in the dataset are:

```{r check-ListingKey}
count(dat, ListingKey, sort = T) %>% filter(n > 1) %>% head(5)
```

For the ListingKey ("17A93590655669644DB4C06") which appears six times in the
dataset, I checked all the six records and found that the `ProsperScore` column
is different among these records. Indeed, this is the only difference exists in
the data and there is not a column which indicates when the `ProsperScore` was
changed. So I prefer to treat these records as problem records and decide to
remove them from the dataset.

```{r, echo=TRUE}
rm_keys <- count(dat, ListingKey, sort = T) %>% filter(n > 1) %$% ListingKey
dat %<>% filter(!(ListingKey %in% rm_keys))
rm(rm_keys)
```

## Check Missing Data

It's an important step to check the missing value in the dataset and decide
whether the missing data will affect the analysis significantly before
performing any analysis. So I calculated the ratio of missing value for each
column of this dataset and printed the columns whose missing data ratio is more
than 30% in the below table:

```{r}
purrr::map_df(colnames(dat), function(x) {
	tibble(variable = x, na_ratio = mean(is.na(dat[x])))
}) %>% arrange(-na_ratio) %>% filter(na_ratio > 0.3)
```

At first it looks like that there are 12 variables whose missing data ratio is
more than 30%, however, after reading the data variables definitions, it turns
out `NA` value of almost all these variables has the specific meaning, which
means they are not really the "missing value". Therefore, I believe that this 
is probably a pretty clean dataset and it's time to play with the data!

# Univariate Plots Section

## Ask the Questions

For the Univariate Plots Section, I choose the following questions for
this project:

1. What's the date period of this dataset?
1. What's the percentage of each ListingCategory in this dataset?
1. What's the percentage of each EmploymentStatus in this dataset?
1. What's the distribution of EmploymentStatusDuration in this dataset?
1. What's the percentage of the Homeowner in the Borrowers?
1. What's the distribution of IncomeRange of the Borrowers?
1. What's the distribution of StatedMonthlyIncome of the Borrowers?
1. What's the distribution of ProsperScore and ProsperRating in this dataset?
1. What's the distribution of BorrowerRate in this dataset?
1. What's the number of the loans in each BorrowerState?
1. What's the percentage of each Occupation in this dataset?
1. What's the distribution of CreditScoreRangeLower and CreditScoreRangeUpper?
1. What's the distribution of LoanOriginalAmount in this dataset?
1. What's the percentage of each LoanStatus in this dataset?
1. What's the percentage of the defaulted loans?
1. what's the distribution of the cycle of these defaulted loans?

## Question: What's the date period of this dataset?

At the first of all, it's useful to check the time period of the dataset:

```{r}
range(dat$ListingCreationDate)
```

It turns out the time period of this dataset is from 2005-11-09 to 2014-03-10.

Now let's see how many listings were created in each day:

```{r}
dat %>% 
	count(Day = as.Date(ListingCreationDate)) %>% 
	ggplot(aes(Day, n)) + 
	geom_col()
```

It's clearly there is a big gap in this dataset, so we should consider that
whether we should ignore the older part of the data or not. For this dataset, I
decided to ignore the older data, and the reasons are as belows:

- the three credit variables `ProsperRating (numeric)`, `ProsperRating (Alpha)`
and `ProsperScore` are only available after 2009-07-13;

- the recent data will be more valuable than the older data to make prediction
in the future;

- the time period of the recent data is more than four years so it's probably
sufficent to build the model.

**So I removed the data before 2009-07-13 from this dataset:**

```{r}
dat <- filter(dat, ListingCreationDate > as.Date("2009-07-13"))
dat %>% count(Day = as.Date(ListingCreationDate)) %>% 
  ggplot(aes(Day, n)) + geom_col()
```

## Question: What's the distribution of the category of the listing?

```{r}
# Recode the ListingCategory to make better visualisations
dat %<>% mutate(
  ListingCategory = fct_recode(
    `ListingCategory (numeric)`,
    "Not Available" = "0",
    "Debt Consolidation" = "1",
    "Home Improvement" = "2",
    "Business" = "3",
    "Personal Loan" = "4",
    "Student Use" = "5",
    "Auto" = "6",
    "Other" = "7",
    "Baby&Adoption" = "8",
    "Boat" = "9",
    "Cosmetic Procedure" = "10",
    "Engagement Ring" = "11",
    "Green Loans" = "12",
    "Household Expenses" = "13",
    "Large Purchases" = "14",
    "Medical/Dental" = "15",
    "Motorcycle" = "16",
    "RV" = "17",
    "Taxes" = "18",
    "Vacation" = "19",
    "Wedding Loans" = "20") %>% 
    fct_infreq() %>% fct_rev()
) 

dat %>% ggplot(aes(ListingCategory)) + geom_bar() + coord_flip()
```

It's interesting to see that the most popular ListingCategory is "Debt
Consolidation", which seems more than the sum of all the other types. The
second ListingCategory is "Other", which provides little effective information
about the usage of the debt.

## Question: What's the percentage of each EmploymentStatus?

```{r}
dat %>%
	mutate(EmploymentStatus = EmploymentStatus %>% 
					 fct_infreq() %>% fct_rev()) %>%
	ggplot(aes(EmploymentStatus)) + 
	geom_bar() + coord_flip()
```

There is no surprise that the majority of the borrowers have a job, because
people with no job have difficulties to obtian the loans.

## Question: What's the distribution of EmploymentStatusDuration?

```{r}
dat %>%
	filter(!is.na(EmploymentStatusDuration)) %>%
	ggplot(aes(EmploymentStatusDuration)) +
	geom_histogram(binwidth = 10, fill = "blue", 
								 color = "black", alpha = 0.6)
```

Summary information of `EmploymentStatusDuration` is: 

```{r}
summary(dat$EmploymentStatusDuration)
```

The figure above shows that the distribution of EmploymentStatusDuration is
positive skewed, and the summary of EmploymentStatusDuration shows that about
one quarter have more than 
`r round(quantile(dat$EmploymentStatusDuration, 0.75, na.rm = T) / 12)` 
years of employment, and about one half have more than 
`r round(quantile(dat$EmploymentStatusDuration, 0.5, na.rm = T) / 12)` 
years of employment.

## Question: What's the percentage of the Homeowner in the Borrowers?

```{r}
dat %>%
	ggplot(aes(IsBorrowerHomeowner)) +
	geom_bar(width = 0.5, fill = "blue", alpha = 0.8)
```

It shows that the homeowner is slightly more than these who don't have a home.

## What's the distribution of IncomeRange of the Borrowers?

```{r}
dat %>%
	ggplot(aes(IncomeRange)) +
	geom_bar() + coord_flip() +
	facet_wrap(~IncomeVerifiable)
```

We can see that the majority have verified income, and most two popular income
ranges are "\$25,000-49,999" and "\$50,000-74,999".

## What's the distribution of StatedMonthlyIncome of the Borrowers?

```{r}
dat %>%
	ggplot(aes(StatedMonthlyIncome)) +
	geom_histogram(binwidth = 100) +
	facet_wrap(~IncomeVerifiable) +
	coord_flip()
```

Wow, the figure above looks bad, most part of it is just blank.
However, we can found an important clue: "some people have a monthly income of
more than \$1500000". 

Because this is a really big number, so the first thought came up to my mind 
is: "Is this income number verified?"

In order to check this, let's filter the records to find who have more than 
\$1500000 monthly income:

```{r}
dat %>% filter(StatedMonthlyIncome > 1500000) %>%
	select(IncomeVerifiable, StatedMonthlyIncome, LoanOriginalAmount)
```

Aha, there is only one person has such a high income but it's not verified. So,
is this a false information or the person is a really wealthy but unwilling to
share his or her information?

After checked every field of this person, I found this person has already
finished 33 trades and has no record of delinquencies. Also, his credit score
looks good.

However, I noticed that the `LoanOriginalAmount` is only "4000", so I am
wondering why a person who can earn more than \$1500000 need to borrow "$4000"? 

This is unreasonable.

So, for now, I would say that this `StatedMonthlyIncome` number is probably not
a true number, even though the person has a good credit history.

Now, let's check the quantile percentage of `StatedMonthlyIncome`:

```{r}
quantile(dat$StatedMonthlyIncome, c(0.01, 0.25, 0.5, 0.75, 0.99, 1.0))
```

And plot the distribution of `StatedMonthlyIncome` without the lowest 1% and the
highest 1%:

```{r}
dat %>%
	filter(StatedMonthlyIncome > quantile(StatedMonthlyIncome, 0.01),
				 StatedMonthlyIncome < quantile(StatedMonthlyIncome, 0.99)) %>%
	ggplot(aes(StatedMonthlyIncome)) +
	geom_histogram(binwidth = 100) +
	facet_wrap(~IncomeVerifiable) +
	coord_flip()
```

Now the visualisation looks more reasonable, which is also a positive skewed
distribution.

## Question: What's the distribution of ProsperScore and ProsperRating in this dataset?

```{r}
dat %>%
	filter(!is.na(ProsperScore)) %>%
	ggplot(aes(factor(as.integer(ProsperScore)))) +
	geom_bar(fill = "blue", alpha = 0.8)
```

The distribution of ProsperScore looks like a normal distribution while the top
3 most frequently ProsperScores are 4, 6, and 8, and the top 2 least frequently
ProsperScores are 1 and 11. However, the data varibale definitions show that the
ProsperScore ranges from 1-10, with 10 being the best, or lowest risk score. So
I decide to double check the ProsperScore field:

```{r}
filter(dat, ProsperScore == 11) %>% count(ProsperScore)
```

**It seems like that there are many rows with the ProsperScore equals 11, but 11
is not a valid score based on the data varibale definitions. So I suspect that
this maybe a data error or something else, which need to be investigated in the
further analysis.**

From the Data variable definitions we know that `ProsperRating (numeric)` and
`ProsperRating (Alpha)` should be consistent with each other, so let's check it:

```{r}
count(dat, `ProsperRating (numeric)`, `ProsperRating (Alpha)`)
```

The result shows they are consistent with each other, so we just need choose one
variable from them, here I choose the `ProsperRating (Alpha)` and rename it to
`ProsperRating`.

```{r}
dat %<>% mutate(ProsperRating = `ProsperRating (Alpha)`)
```

`ProsperRating` should have some strong relationship with `ProsperScore`, but
first let's check the `ProsperScore` when `ProsperRating` is `NA`：

```{r}
dat %>%
	count(is.na(ProsperScore), is.na(ProsperRating))
```

The result shows that when `ProsperRating` is `NA`, the `ProsperScore` is also
`NA`, and vice versa. Pretty good!

Now let's check the relationship between `ProsperRating` and `ProsperScore`:

```{r}
dat %>% 
	filter(!is.na(ProsperRating)) %>%
  count(ProsperRating, ProsperScore) %>%
	ggplot(aes(ProsperRating, ProsperScore)) + 
  geom_tile(aes(fill = n), alpha = 0.9) +
  scale_fill_distiller(
    type = "seq", palette = "OrRd", direction = 1
  ) + theme_minimal()
```

The above heatmap shows that there is a positive relationship between 
`ProsperScore` and `ProsperRating`, however, in each category of the 
`ProsperRating`, there is obvious variation in `ProsperScore`, 
which means `ProsperScore` can provide addtional information along with 
`ProsperRating`, so both variables will be kept for future use.

## What's the distribution of BorrowerRate in this dataset?

```{r}
dat %>%
  ggplot(aes(BorrowerRate)) +
  geom_histogram(
    binwidth = 0.01, color = "darkred",
    fill = "grey", alpha = 0.9)
```

The above histgoram plot shows that the distribution of BorrowerRate roughly 
follows a normal distribution, however there is a pinnacle at 0.32, which is a 
rather high BorrowerRate.

## What's the number of the loans in each BorrowerState?

```{r, fig.asp=1}
dat %>% ggplot(aes(BorrowerState %>% fct_infreq() %>% fct_rev())) + 
  geom_bar() + coord_flip() + labs(x = "BorrowerState")
```

The above plot shows that the number of prosper loans in CA is far more than 
any other states, followed by NY, TX, and FL, however, some states such as 
WY, AK, or VT have only a few loans.

## What's the percentage of each Occupation in this dataset?

```{r, fig.asp=1}
dat %>% ggplot(aes(Occupation %>% fct_infreq() %>% fct_rev())) + 
  geom_bar() + coord_flip() + labs(x = "Occupation")
```

The above plot clearly shows that most borrowers were either unwilling to provide their 
occupation information or have unusual occupation, because there are a lot 
of "Other" value in the `Occupation` field.

## What's the distribution of CreditScoreRangeLower and CreditScoreRangeUpper?

```{r, fig.asp=1}
gridExtra::grid.arrange(
  ggplot(dat, aes(CreditScoreRangeLower)) + geom_bar(),
  ggplot(dat, aes(CreditScoreRangeUpper)) + geom_bar(),
  nrow = 2
)
```

The distribution of CreditScoreRangeLower and CreditScoreRangeUpper are quite 
similar and both of them follow a slightly positive skewed distribution.

## What's the distribution of LoanOriginalAmount in this dataset?

```{r}
dat %>% ggplot(aes(LoanOriginalAmount)) + geom_histogram(binwidth = 1000)
```

It's interesting to see that there are several peaks in the histogram: 
4000, 10000, 15000, 20000, 25000, which is reasonable in the real world.

## Question: What's the percentage of each LoanStatus?

```{r echo=FALSE}
dat %>% 
	mutate(LoanStatus = LoanStatus %>% fct_infreq() %>% fct_rev()) %>%
	ggplot(aes(LoanStatus)) + geom_bar() + coord_flip()
```

We can see that only a little fraction of the loans are in the "Past Due"
status, and generally the longer the past due days, the less they are.

However, the number of the "Chargedoff" loans is 
`r sum(dat$LoanStatus == 'Chargedoff')` and the number of the "Defaulted" loans 
is `r sum(dat$LoanStatus == 'Defaulted')`, which means the default risk is high 
and should be addressed by the investors (Loans with a "Chargedoff" or 
"Defaulted" status are about 
`r paste0(round(100*mean(dat$LoanStatus %in% c("Chargedoff","Defaulted")),2),"%")` 
of the total loans).

When I read the data variable definitions, I noticed that there is a column
called `LoanFirstDefaultedCycleNumber` which indicates the cycle when the loan
was charged off, and if the value of this column is `NA`, it means that the loan
hasn't been charged off. So I want to check whether this column is consistent
with the `LoanStatus` column:

```{r}
dat %>%
	filter(LoanStatus == "Chargedoff", 
				 is.na(LoanFirstDefaultedCycleNumber)) %>%
	select(LoanStatus, LoanFirstDefaultedCycleNumber)
```

There are seven rows which have a "Chargedoff" `LoanStatus` while have a `NA`
value in `LoanFirstDefaultedCycleNumber` column, so these records are not
consistent.

The `LoanStatus` variable also has a possible value "Defaulted", and records
with a "Defaulted" `LoanStatus` also should not have a `NA` value in
`LoanFirstDefaultedCycleNumber` column, let's check this:

```{r}
dat %>%
	filter(LoanStatus == "Defaulted", 
				 is.na(LoanFirstDefaultedCycleNumber)) %>%
	select(LoanStatus, LoanFirstDefaultedCycleNumber)
```

There are 100 inconsistent records, and I suppose the reason that
`LoanFirstDefaultedCycleNumber` has a `NA` value for `Chargedoff` and
`Defaulted` loans is just because the defaulted cycle number haven't been
recorded correctly, so even the `LoanFirstDefaultedCycleNumber` is `NA` these
loans are still defaulted loans, which is similar for the `Chargedoff` loans.

So, for this project, I decided to define the defaulted loans as those 
meet the below criterias:

1. `LoanStatus` is either `Chargedoff` or `Defaulted`; or,
2. `LoanFirstDefaultedCycleNumber` is not `NA` and `LoanStatus` is `Completed`.

Also, I will define the non-defaulted loans whose `LoanStatus` are `Completed`
and `LoanFirstDefaultedCycleNumber` is `NA`, **which means I only use the**
**records whose `LoanStatus` is `Chargedoff` or `Defaulted` or `Completed`.**

```{r}
dat %<>%
	filter(LoanStatus %in% c('Chargedoff', 'Defaulted', 'Completed'))
```

Now, there are `r nrow(dat)` records left in the dataset.

## Question: What's the percentage of the defaulted loans?

First, let's create a new variable to indicate whether a loan is defaulted or
not:

```{r, echo=TRUE}
dat %<>%
	mutate(is_defaulted = as.factor(if_else(
		(LoanStatus %in% c("Chargedoff", "Defaulted")) | 
			(LoanStatus == "Completed" & !is.na(LoanFirstDefaultedCycleNumber)),
		"Defaulted", "Not Defaulted"
	)))
```


Then, we can create a bar plot to show the percentage of the defaulted loans:

```{r}
dat %>%
	ggplot(aes(is_defaulted)) + 
	geom_bar(width = 0.5, fill = "blue", alpha = 0.8)
```

## Question: What's the distribution of the cycle of these defaulted loans?

Furthermore, the distribution of the cycle of these defaulted loans may reveal
more information:

```{r}
dat %>%
	filter(!is.na(LoanFirstDefaultedCycleNumber)) %>%
	ggplot(aes(LoanFirstDefaultedCycleNumber)) +
	geom_bar(fill = "blue", alpha = 0.8)
```

It's interesting to find that if default didn't occur at the beginning cycle,
then it's unlikely to occur in the following three cycles. Interesting.

# Univariate Analysis

## What is the structure of your dataset?

There are 113,937 loans with 81 variables including *loan amount*, 
*borrower rate (or interest rate)*, *current loan status*, *borrower income*, 
*borrower employment status*, *borrower credit history*, and 
*the latest payment information* in this dataset.

There are so many variables in this dataset so it is difficult to explore each
variable one by one. Instead, it woule be more appropriate to adopt a
problem-oriented approach to explore this dataset. So I choose this
problem-oriented approach to finish this project.

## What is/are the main feature(s) of interest in your dataset?

The main features of interest are `LoanFirstDefaultedCycleNumber`
and `LoanStatus`, because for the loan business, the most important concern is
whether and when the loan will default. (i.e. credit/default risk)

For the creditor/lender, what they really care about is whether the borrower
will default on a specific loan and when the borrower will most likely default.

**For this project, I will only focus on whether the borrower will default.**
So I created a new feature with the name "is_defaulted" and 
choose it as the main feature of interest.

## What other features in the dataset will help support your investigation?

Of all the features in this dateset, I will use `ProsperScore`, `ProsperRating`,
`CreditScoreRangeLower`, `CreditScoreRangeUpper` along with some other features
(more details in "Bivariate Plots Section"") to predict whether 
the a specific loan will default or not.

## Did you create any new variables from existing variables in the dataset?

Yes, I create `is_defaulted` from `LoanFirstDefaultedCycleNumber` and
`LoanStatus`, and I modified some variables (changing the factor levels or
reorder the factor or rename the variable) in order to make better
visualisations.

In addition, in the Bivariate Plots (Analysis) Section and Multivariate Plots
(Analysis) Section, I will create more new variables to get better understanding
of the dataset and build better models.

## Of the features you investigated, were there any unusual distributions?

Yes, such as `EmploymentStatusDuration` and `LoanFirstDefaultedCycleNumber`,
they don't follow normal distribution and both of them are positive skewed.

**Did you perform any operations on the data to tidy, adjust,**
**or change the form of the data? If so, why did you do this?**

Yes, as I said before, I changed the form of some variables (changing the 
factor levels or reorder the factor) in order to make better visualisations 
and I checked the time period of this dataset to choose a subset of the data. 
Also, I checked the consistency for some related features such as `LoanStatus` 
and `LoanFirstDefaultedCycleNumber`.

After checked the dataset, I decided to use only the records whose `LoanStatus`
is `Chargedoff` or `Defaulted` or `Completed`. As a result, only about 
`r paste0(round(nrow(dat) / 113937, 2) * 100, "%")` of the original dataset 
were chosen to perform further analysis and build the model.

# Bivariate Plots Section

In this section, I will mainly explore the relationships between the target
feature and the support features, inculding the relationship between the target
feature and a single support feature and the relationship between the target
feature and multiple support features. In addition, the relationships between
support features will be explored as well to obtain more insights and address
the multicollinearity problems in the model stage.

## Define the target feature

The first and the most important step is defining the target feature properly,
which was already done in the pervious section.

## Determine the input features

Generally, whether a loan will default has strong relationship with the credit
status of the borrower. In this dataset, there are two main variables which
indicate the credit of the borrower: `ProsperRating` and `ProsperScore`.

Now, it's time to visualise the relationship between default status with
`ProsperScore` and `ProsperRating`, first let's check the `ProsperRating`:

```{r}
dat %>% 
  ggplot(aes(ProsperRating, fill = is_defaulted)) + 
  geom_bar(position = "fill")
```

It's no surprise that the percent of the defaulted loan has an inverse
realitionship with `ProsperRating`, that is, better `ProsperRating` means lower
default rate.

Second, let's check the `ProperScore`:

```{r}
dat %>% 
  ggplot(aes(ProsperScore, fill = is_defaulted)) + 
  geom_bar(position = "fill")
```

Form this figure we can notice that the percent of the defaulted loan doesn't
have an exact inverse relationship with `ProsperScore`, there are an exception
when `ProsperScore` arounds 4.

Besides the `ProsperScore` and `ProsperRating`, there are other features may
have relationships with the defaulted rate, a rigorous way to identify all the
potential useful features is digging into each feature one by one, or divided
all the feature into different groups based on their type and use some screening
skills to pick out the most useful features. However, I won't do this here in 
order to avoid making this report too long to read. Instead, here I will quickly
create a correlation matrix plot for the numeric variables to see whether there
are highly correlated features in the dataset.

```{r, fig.asp=1.2, fig.width=8}
GGally::ggcorr(
  dat, geom = "circle", min_size = 0, max_size = 5,
  hjust = 1, size = 3, color = "grey50", layout.exp = 6
)
```

From the above plot we see that some features have a very high correlation 
between each other, and features with high correlation probably have a lot of
inforamtion in common and may give rise to multicollinearity problems in the 
model stage.

In this project, I used the correlation matrix plot as a reference and then 
choose the features (only a few features here, not the whole set for 
real life project) which may have relationships with defaulted rate base
on the domain knowledge:

- `BorrowerRate`: 
    The higher the `BorrowerRate`, maybe the higher the defaulted probability;
- `BorrowerState`: 
    The defaulted rate may be different in each State 
    due to their different economic condition;
- `EmploymentStatus` and `EmploymentStatusDuration`: 
    The borrower without a job may be more likely to default;
- `Occupation`: 
    Different occupation may have different defaulted probability;
- `IsBorrowerHomeowner`: 
    The defaulted probability may be different between those 
    who own a home and those who don't;
- `CreditScoreRangeLower` and `CreditScoreRangeUpper`: 
    Provide similar information with `ProsperScore` and `ProsperRating`;
- `IncomeRange`: 
    Borrowers with different IncomeRange may have different default probability;
- `IncomeVerifiable`: 
    Borrowers with verified income may have lower defaulted probability;
- `LoanOriginalAmount`: 
    Loan with different Original amount may have different default probability;
- `ListingCategory`: 
    Loan of different ListingCategory may have different default probability.

There are two variable types here: categorical variables and numeric variables.

For categorical variables, I will create a bar plot to show the default rate
for each level (use a custom defined function named `bar_plot`). And for the 
numeric variables I will create a boxplot along with the violin plot to show 
the distribution of the variable separately for default loans and non-default 
loans (use a custom defined function named `violin_boxplot`).

```{r}
bar_plot <- function(df, col, facet_ncol = 4) {
	col <- enquo(col)
	category_num <- df %>% count(!!col) %>% nrow()
	# print(category_num)
	if (category_num < 24) {
		df %>% 
			mutate(col = fct_reorder(
				(!!col), is_defaulted, function(x){mean(x == "Yes")})) %>% 
			ggplot(aes(is_defaulted, fill = is_defaulted)) + geom_bar() + 
			facet_wrap(~col, scales = "free_y", ncol = facet_ncol) + 
			theme_bw(base_family = "Times New Roman") +
			theme(legend.position = "bottom", 
						plot.title = element_text(hjust = 0.5)) + 
			labs(title = col)
	} else {
		df %>% 
			group_by(!!col) %>% 
			summarise(default_ratio = mean(is_defaulted == "Defaulted")) %>% 
			mutate(col = fct_reorder(as.factor(!!col), -default_ratio)) %>% 
			ggplot(aes(col, default_ratio)) + 
			theme_bw(base_family = "Times New Roman") +
			geom_col(width = 0.8, fill = "red", alpha = 0.6) + 
			xlab(col) + coord_flip()
	}
}

violin_boxplot <- function(df, col) {
	col <- enquo(col)
	df %>% mutate(x = (!!col)) %>%
		filter(x > quantile(x, 0.01, na.rm = T), 
					 x < quantile(x, 0.99, na.rm = T)) %>%
		ggplot(aes(is_defaulted, x)) + geom_violin() + 
		geom_boxplot(fill = "blue", alpha = 0.3) + 
		ylab(col) + theme_bw(base_family = "Times New Roman")
}
```

### Categorical input variables

#### BorrowerState

```{r, fig.asp=1}
bar_plot(dat, BorrowerState)
```

It seems like there is obvious difference among different states. Also, let's
create a map to see whether there is some spatial pattern for default rate:

```{r}
states <- map_data("state")
data(state)
state_name_abb <- tibble(state_abb = state.abb, 
												 state_name = stringi::stri_trans_tolower(state.name))
states <- left_join(states, state_name_abb, by = c("region" = "state_name"))

dat %>% group_by(BorrowerState) %>%
	summarise(DefaultRatio = mean(is_defaulted == "Defaulted")) %>%
	mutate(BorrowerState = as.character(BorrowerState)) %>%
	right_join(states, by = c("BorrowerState" = "state_abb")) %>%
	ggplot(aes(long, lat)) + coord_map() + theme_void() +
		geom_polygon(aes(group = group, fill = DefaultRatio), alpha = 0.9) +
	scale_fill_gradient(low = "yellow", high = "red")
```


#### EmploymentStatus

```{r, fig.asp=1.3}
gridExtra::grid.arrange(
  bar_plot(dat, EmploymentStatus, facet_ncol = 1),
  ggplot(dat, aes(fct_rev(EmploymentStatus))) + 
    geom_bar(fill = "red", alpha = 0.6, width = 0.8) + 
    coord_flip() + theme_bw(base_family = "Georgia") +
    labs(x = ""),
  nrow = 1
)
```

We can see that the default ratio of the borrowers whose `EmploymentStatus` is
either "Not employed" or "Other" is higher than those who are employed. Also,
we should noticed that the number of loans has a high variance among different
`EmploymentStatus` groups.

#### Occupation

```{r, fig.asp=1.3}
bar_plot(dat, Occupation)
```

We can see that different occupation has different default ratio, and the 
default ratio of the sophomore college student is as high as more than 80%.

#### IsBorrowerHomeowner

```{r}
bar_plot(dat, IsBorrowerHomeowner)
```

We can see that the defualt ratio of the borrowers who have a home is slightly
lower than those who don't have a home.

#### IncomeRange

```{r}
bar_plot(dat, IncomeRange)
```

We can clearly see that higher income means lower default ratio.

#### IncomeVerifiable

```{r}
bar_plot(dat, IncomeVerifiable)
```

We can see that the defualt ratio of the borrowers whose income are verified is
slightly lower than those whose income are not verified.

#### ListingCategory

```{r, fig.asp=1}
bar_plot(dat, ListingCategory)
```

We can see that different `ListingCategory` have quite different default ratios.

### Numeric input variables

#### BorrowerRate

```{r}
violin_boxplot(dat, BorrowerRate)
```

We can see that the higher the `BorrowerRate`, the higher the default ratio.

#### EmploymentStatusDuration

```{r}
violin_boxplot(dat, EmploymentStatusDuration)
```

We can see that `EmploymentStatusDuration` has no clear relatioship with default ratio.

#### CreditScoreRangeLower

```{r}
violin_boxplot(dat, CreditScoreRangeLower)
```

#### CreditScoreRangeUpper

```{r}
violin_boxplot(dat, CreditScoreRangeUpper)
```

We can see that the lower the `CreditScoreRangeLower` and
`CreditScoreRangeUpper`, the higher the default ratio.

#### LoanOriginalAmount

```{r}
violin_boxplot(dat, LoanOriginalAmount)
```

We can see that loans with smaller `LoanOriginalAmount` tend to be more likely
to default. Interesting.

# Bivariate Analysis

## Talk about some of the relationships you observed in the investigation. 

I explored the relationships between the target variable `is_defaulted` and 14
predictors, including 8 categotical variables and 6 numeric variables.

**How did the feature(s) of interest vary with other features in the dataset?**

The feature of interest vary with a lot of other features in this dataset. For
the 14 variables that I chose in this project, almost all of them have some
relationship with the target variable except for `EmploymentStatusDuration`
based on the bar plots and violon-boxplot. For example, the plots above show
that the default rate has a clear relatioship with `ProsperScore`,
`ProsperRating`, `EmploymentStatus`, `Occupation`, `IncomeRange`, and
`BorrowerRate`.

## Did you observe any interesting relationships between the other features?

Yes, there is an interesting relatioship between `IncomeRange` and
`IncomeVerifiable`:

```{r}
dat %>% 
	mutate(IncomeRange = relevel(IncomeRange, "Not employed")) %>% 
	ggplot(aes(IncomeVerifiable, fill = IncomeVerifiable)) + 
	geom_bar(alpha = 0.8) + 
	facet_wrap(~IncomeRange, scales = "free_y", ncol = 4) + 
	theme(legend.position = c(0.88, 0.25), 
				plot.title = element_text(hjust = 0.5)) + 
	labs(title = "IncomeRange", x = "") + 
	scale_fill_manual(values = c("green", "red"))
```

It shows clearly that most of the borrowers who are either not employed or have
zero income (maybe just unwilling to provide the income status) have unverified
income, also according to the relationship between `is_defaulted` and
`IncomeVerifiable` we see that borrowers with unverified income have a higher
default rate. So we can use these variables to predict the default rate
for new loans.

## What was the strongest relationship you found?

The strongest relationship I found is between the `is_defaulted` and
`ProsperRating`, better `ProsperRating` means lower default rate without
exception, which means the `ProsperRating` is a very important
factor to be considered when investing in a prosper loan.

Also, I found an interesting pattern for the college student group:

```{r}
dat %>% 
	filter(stringi::stri_startswith_fixed(Occupation, "Student")) %>%
	group_by(Occupation) %>% 
	summarise(n = n(), defaulted_ratio = mean(is_defaulted == "Defaulted")) %>% 
	mutate(Occupation = fct_relevel(
		Occupation,
		"Student - College Freshman",
		"Student - College Sophomore",
		"Student - College Junior",
		"Student - College Senior",
		"Student - College Graduate Student",
		"Student - Technical School",
		"Student - Community College"
	)) %>%
	ggplot(aes(Occupation, defaulted_ratio)) +
	geom_col(fill = "red", alpha = 0.6) + 
	geom_label(aes(Occupation, defaulted_ratio, label = n), nudge_y = 0.03) +
	coord_flip()
```

The above figure shows that for the College Student: higher grade means lower
default rate, with exception for Sophomore Student. The labels in the figure
stand for the sample size for each group, which means about 10 out of the 12
Sophomore Students were defaulted. This is a really high ratio.

# Multivariate Plots Section

From previous sections we found that default rate has a strong relationship
with `ProsperRating` and `ProsperScore`, also the default rate should change
over time, so I decided to create a time series plot for both default rate
and `ProsperScore`:

```{r Multivariate_Plots}
dat %>%
	group_by(
		Time = strftime(ClosedDate, "%Y-%m") %>% 
			paste("15", sep = "-") %>% 
			as.Date()
		) %>%
	summarise(DefaultRatio = mean(is_defaulted == "Defaulted"),
						ProsperScore = mean(ProsperScore, na.rm = T)) %>%
	gather(key, value, -Time) %>%
	ggplot(aes(Time, value)) + geom_line(aes(Time, value), linetype = 2) + 
	geom_point(size = 0.2) + geom_smooth() + 
	facet_wrap(~key, scales = "free_y", ncol = 1)
```

It clearly shows that as the time goes by the ProsperScore became lower and 
lower but the DefaultRatio became higher and higher.

## Building the Model

The model part has three steps:

1. Data Preparation;
2. Model Training;
3. Model Evaluation.

### Data Preparation

The target variable is `is_defaulted` and I choose 13 features as the input
variables of the model. 

```{r}
ds <- dat %>%
	arrange(ListingCreationDate) %>%
	select(
		is_defaulted,
		ProsperRating,
		ProsperScore,
		BorrowerState,
		EmploymentStatus,
		Occupation,
		IsBorrowerHomeowner,
		IncomeRange,
		IncomeVerifiable,
		ListingCategory,
		BorrowerRate,
		CreditScoreRangeLower,
		CreditScoreRangeUpper,
		LoanOriginalAmount
	) %>% na.omit() %>%
  mutate(is_defaulted = if_else(is_defaulted == "Defaulted", 1, 0))
```

After the data preprocessing of all the previous sections, now there are 
`r nrow(ds)` rows left in the dataset to build the model.

In order to evaluate the performance of the model, we need to split the dataset
into two parts: train dataset and test dataset. Here I used 80% of the data to
train the model, and used the left 20% of the data to evaluate the model.


```{r}
set.seed(123)
train_rows <- sample.int(nrow(ds), 0.8 * nrow(ds), replace = FALSE)
train_set <- ds[train_rows, ]
test_set <- ds[-train_rows, ]
```

### Model Training

In this project I tried Logistic Regression and Support Vector Machine to build
the model:

**Logistic Regression Model**

```{r}
# the training process is time consuming, so I save the model to reuse.
if (file.exists("data/model_glm.rda")) {
	load("data/model_glm.rda")
} else {
	print("Change the 'eval' chunk option of 'model_glm' from FALSE to TRUE.")
}
```

> Notice: Change the `eval` option to `TRUE` if the model need to be updated.

```{r model_glm, echo=TRUE, eval=FALSE}
model_glm <- glm(formula = is_defaulted ~ ., data = train_set, family = binomial)
save(model_glm, file = "data/model_glm.rda")
```

**Support Vector Machine**

```{r}
# the training process is time consuming, so I save the model to reuse.
if (file.exists("data/model_svm.rda")) {
	load("data/model_svm.rda")
} else {
	print("Change the 'eval' chunk option of 'model_svm' from FALSE to TRUE.")
}
```

> Notice: Change the `eval` option to `TRUE` if the model need to be updated.

```{r model_svm, echo=TRUE, eval=FALSE}
model_svm <- svm(is_defaulted ~ ., data = train_set, probability = TRUE)
save(model_svm, file = "data/model_svm.rda")
```

### Model Visualisation and Evaluation

```{r, warning=FALSE}
glm_result_test <- dplyr::mutate(
	test_set,
	pred = as.numeric(predict(model_glm, test_set, type = "response"))
) %>% select(is_defaulted, pred) %>%
  mutate(is_defaulted = as.factor(is_defaulted))
```

```{r}
svm_result_test <- dplyr::mutate(
	test_set,
	pred = attr(predict(model_svm, test_set, probability = TRUE), 
							"probabilities") %>% as.tibble() %$% Defaulted
) %>% select(is_defaulted, pred) %>%
  mutate(is_defaulted = as.factor(is_defaulted))
```


```{r}
# function to visualise the model performance
plot_result <- function(result) {
  result %>%
    mutate(
    is_defaulted = if_else(is_defaulted == 1, "Defaulted", "Not Defaulted")
  ) %>%
  ggplot() +
		geom_jitter(aes(is_defaulted, pred, color = is_defaulted), 
								width = 0.05, size = 0.01, alpha = 0.5) +
		theme_bw() + theme(legend.position = "none")
}
```

First, let's visualise the performance of the Logistic Regression model:

```{r}
plot_result(glm_result_test)
```

Then, let's visualise the performance of the Support Vector Machine model:

```{r}
plot_result(svm_result_test)
```

We can see both the logistic regression model and support vector machine model
are not good, no matter which threshold we choose to decide whether a loan will
default or not, the result is not satisfied.

For the classification task, the most relevant performance metrics are
Precision, Recall, and F1 Score:

- Precision: (True Positive) / (True Positive + False Positive)
- Recall: (True Positive) / (True Positive + False Negative)
- F1 Score: Harmonic Mean of Precision and Recall.

One thing that we should keep in mind is that generally we need to make tradeoff
between Precision and Recall, so whether we should choose the model with high
precision or high recall depends on the real problem and our target.

For example, if we want to make the recall ratio higher, we can choose a lower
threshold for the model, in fact, if we simply predict "Positive" for each sample,
the recall ratio will equal 1. And if we want to make the precision ratio
higher, we can choose a higher threshold for the model. In order to use a single
ratio to compare different ratio, we can use the F1 Score:

$$F1\;Score = \frac{2\cdot Precision\cdot Recall}{Precision + Recall}$$

```{r}
# function to calculate precision, recall, and F1 Score
model_metrics <- function(cutoff, result) {
	pred <- result["pred"]
	cmatrix <- table(result$is_defaulted, pred > cutoff)
	# print("Confusion Matrix:")
	# print(cmatrix)
	precision <- cmatrix[2,2] / (cmatrix[2,2] + cmatrix[1,2])
	recall <- cmatrix[2,2] / (cmatrix[2,2] + cmatrix[2,1])
	f1_score <- 2 * precision * recall / (precision + recall)
	# print(paste("Precision is:", as.character(precision)))
	# print(paste("Recall is:", as.character(recall)))
	# print(paste("F1 Score is:", as.character(f_score)))
	metrics <- tibble::tibble(
		cutoff = cutoff,
		precision = precision,
		recall = recall,
		f1_score = f1_score
		)
	return(metrics)
}

# compute precision, recall, and F1 Score for a lot of thresholds
compute_pc_table <- function(result) {
	thresholds <- seq(min(result$pred), max(result$pred), by = 0.005)
	purrr::map(thresholds, model_metrics, result = result) %>% bind_rows()
}

# F1 Score vs. Thresholds
plot_f1_score <- function(pc_table) {
	ggplot(pc_table, aes(cutoff, f1_score)) + 
		geom_point() + 
		geom_smooth(method = "loess") +
		facet_wrap(~model) + 
		theme_bw()
}

# Precison-Recall Curve
plot_pc <- function(pc_table) {
	ggplot(pc_table, aes(recall, precision)) + 
		geom_point() +
		geom_smooth(method = "loess") +
		facet_wrap(~model) + 
		theme_bw()
}
```


```{r}
pc_table_glm <- compute_pc_table(glm_result_test) %>% mutate(model = "glm")
pc_table_svm <- compute_pc_table(svm_result_test) %>% mutate(model = "svm")
pc_table <- bind_rows(pc_table_glm, pc_table_svm)
```

First, let's plot the F1 Score vs. Thresholds for each model we built:

```{r}
plot_f1_score(pc_table)
```

Then, let's plot the Precison-Recall Curve for each model we built:

```{r}
plot_pc(pc_table)
```

Finally, let plot the ROC curve and calculate the AUC ratio:

ROC curve of Logistic Regression Model is:

```{r}
plot(roc(glm_result_test$pred, glm_result_test$is_defaulted))
```

and the AUC ratio for Logistic Regression Model is 
`r auc(roc(glm_result_test$pred, glm_result_test$is_defaulted))`.

ROC curve for Support Vector Machine is:

```{r}
plot(roc(svm_result_test$pred, svm_result_test$is_defaulted))
```

and the AUC ratio for Support Vector Machine is 
`r auc(roc(svm_result_test$pred, svm_result_test$is_defaulted))`.

# Multivariate Analysis

## Talk about some of the relationships you observed in this part of the investigation.

From the plot in the Multivariate Plots Section we noticed that as the time goes
by, the ProsperScore became lower and lower and the DefaultRatio became higher
and higher, which means the defaulted risk of the Prosper Loan Business should
be addressed by the investors and they should be very careful when make
investments on Prosper Loan.

## Were there any interesting or surprising interactions between features?

The most interesting pattern I found is the default rate within the student
group. Generally, higher grade means lower default rate, however, this is not
true for the Sophomore Student: there were about 10 out of 12 Sophomore Students
who were defaulted, which was a really high ratio.

## OPTIONAL: Did you create any models with your dataset? 

**Discuss the strengths and limitations of your model.**

Yes, I created two models with this dataset: Logistic Regression and Support
Vector Machine.

From the model evaluation part we can see that both of the models have some
predictive power for default rate, however, this predictive power is weak.
The reason may be that I haven't performed the parameter tuning for these model,
also I think that feature engineering is also very important if we want to build 
a better model.

------

# Final Plots and Summary

## Plot One

```{r Plot_One, echo=FALSE}
dat %>% mutate(
		IncomeRange = fct_reorder(
			IncomeRange, 
			is_defaulted, 
			function(x) { 
				mean(x == "Yes")
			})) %>%
	ggplot(aes(is_defaulted, fill = is_defaulted)) + 
	geom_bar() + facet_wrap( ~ IncomeRange, scales = "free_y", ncol = 4) +
	theme_bw(base_family = "Georgia") +
	theme(legend.position = c(0.88, 0.3),
				plot.title = element_text(hjust = 0.5)) +
	labs(x = "Is Defaulted or Not", y = "Number of Loans", 
			 fill = "Is Defaulted or Not",
			 title = "Defaulted Loans for Different IncomeRange Groups")
```

## Description One

The plot one indicates a strong relationship between IncomeRange and the default
ratio. For the borrwoers who have no income or are not employed, the default
ratio is very high. Also, there is an exact inverse relationship between income
and the default ratio: the higher the income, the lower the default ratio.

## Plot Two

```{r Plot_Two, echo=FALSE}
dat %>% 
	filter(stringi::stri_startswith_fixed(Occupation, "Student")) %>%
	group_by(Occupation) %>% 
	summarise(n = n(), default_ratio = mean(is_defaulted == "Defaulted")) %>% 
	mutate(Occupation = fct_relevel(
		Occupation,
		"Student - College Freshman",
		"Student - College Sophomore",
		"Student - College Junior",
		"Student - College Senior",
		"Student - College Graduate Student",
		"Student - Technical School",
		"Student - Community College"
	)) %>%
	ggplot(aes(Occupation, default_ratio)) +
	geom_col(fill = "red", alpha = 0.6) + 
	geom_label(aes(Occupation, default_ratio, label = n), nudge_y = 0.03) +
	theme_bw(base_family = "Georgia") + coord_flip() +
	scale_y_continuous(breaks = seq(0, 1, 0.1)) +
	labs(subtitle = "Interesting Default Ratio Pattern among Different Student Groups",
			 y = "Default Ratio", x = "") +
	theme(plot.subtitle = element_text(hjust = 0.5))
```

## Description Two

The plot two reveals a very interesting pattern of the default rate among 
different college student groups: the sophomore college students have the 
highest default rate than any other college student grouop, followed by the 
freshman, junior, senior and Graduate. 

Additionally, the number of loans has an inverse trend.

We can also notice that students from community college have higher default
ratio than those from technical school, however, the sample number of the
students from technical school is just one, which means it has no statistically
meaning.

## Plot Three

```{r Plot_Three, echo=FALSE}
dat %>%
	group_by(
		Time = strftime(ClosedDate, "%Y-%m") %>% 
			paste("15", sep = "-") %>% 
			as.Date()
		) %>%
	summarise(`Default Ratio` = mean(is_defaulted == "Defaulted"),
						`Average ProsperScore` = mean(ProsperScore, na.rm = T)) %>%
	gather(key, value, -Time) %>%
	ggplot(aes(Time, value)) + geom_line(aes(Time, value), linetype = 2) + 
	geom_point(size = 0.2) + geom_smooth() + 
	facet_wrap(~key, scales = "free_y", ncol = 1) +
	theme_bw(base_family = "Georgia") +
	labs(title = "Default Ratio and ProsperScore Change over Time",
			 x = "Closed Date of the Loan", y = "Default Ratio vs. Prosper Score") +
	theme(plot.title = element_text(hjust = 0.5))
```

## Description Three

The plot three shows that there is a inverse relationship between 
ProsperScore and Default Rate, which is no surprise. 
Also, it shows that how these two indicators change over time. 

An important finding is that as time goes by:

- the credit score of the borrowers keep going down;
- the default rate increased quickly from 2010 to 2011;
- the default rate maintained a higher ratio after 2011.

The conclusion is that the risk of the Prosper Loan Business had been higher
than the days when it started, so investors should be prudent and thoughtful
when they make new investments on this platform.

------

# Reflection

After performed the exploration on this dataset, I found a lot of insights about
the loan business. Most of the insights are just the confirmation of the
knowledge we already have: such as better credit score and higher income lead to
lower default ratio, and different states or different occupations have
different default ratio, and so on.

Besides, I have found two interesting insights I haven't known before:

1. For college students, after they entered into the college, more and more of
them may choose to borrow money as time went by, and the higher the grade, the
lower the default ratio, with exception for the sophomore students. So why
the sophomore students have such a high default ratio? This is an interesting
question and I would like to do some research about it in the future.

2. The risk of the Prosper Loan Business has become higher than the days when it
started, which is evidenced by the higher default rate and lower ProsperScore 
than before.

Besides the exploration analysis and visualisation on this dataset, I also built
two models to try to predict the default possibility of a certain loan based on
the features of the borrower as well as the features of the loan itself.

However, the performance of the models were not good and the reasons are:

1. Feature selection and feature engineering were not performed yet, which play
an important role for machine learning models. In fact, the selected features
may have some collinearity, because we got the message `prediction from a
rank-deficient fit may be misleading` from `predict.lm(object, newdata, se.fit,
scale = 1, ...` (perhaps we can also try deep learning to make the feature
engineering process automatically);

1. Model parameters were not tuning to achieve the potential highest performance
of the model, and other models such as GBRT and Random Forest should be tested
as well.

Also, there are sereval issues that we should addressed in the future:

1. In the model training part, I randomly chose 80% of the data as train data
and use the left 20% of the data as test data, which seems reasonably at
first. However, if we think about our target problem carefully, we may notice
that once we have a model we need use it to predict the future outcome of the
currently unfinished loans. So, we are forecasting the future. And for this kind
of the problem, the test daatset should be the recent part and the train dataset
should be the older part of the whole dataset. We can also use the moving time
window technique to generate the train and test dataset;

1. Predict whether a loan will default or not is really a difficult task 
because this is predicting the future, and the future is definitely full 
of the uncertainty. For example, even a borrower with low income may repay 
his(her) debt on time if he has be abitily to clear the debt when it matures, 
conversely, even a borrower with high income may defualt on his(her) loan 
if he(she) has some financial difficulties when the debt matures. 
In addition, even for the same borrower, the default possibility may be 
different for different loans just because the consequence of default 
are different for different loans (ie. mortgage loan vs. credit card).

1. From final plot 3 we see that the default ratio for all the loans changes
over time, which means besides the features of the borrower and the features of
the loan itself there are some other time-related features have impact on the
defualt possibility: the Macroeconomic Condition and Currency Policy of
differnent time periods. 
Therefore, if we really want to build a model which can be used in the real 
world, we should take the Macroeconomic Condition and Currency Policy into 
consideration as well as the features about the borrowers and the loan.
